---
title: "bayes proj"
author: "Claire Morrison"
date: "4/26/2022"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(psych)
library(lme4)
library(lmerTest)
library(dplyr)
library(rethinking)
library(tidyverse)
library(ggridges)
library(teamcolors)
```


```{r}
nba<- fread("/Users/clairemorrison/Documents/Classes/bayes//All_NBA_Data.csv", header=T, data.table=F)
colnames(nba)
nrow(nba)
length(unique(nba$game_id)) ### 1230 games
head(nba)



ppy<- fread("/Users/clairemorrison/Documents/Classes/bayes//pts_py.csv", header=T, data.table=F)
ggplot(data = ppy, aes(x = Season , y = PTS)) +
    geom_point()+
  geom_line()


ggplot(ppy, aes(x=Season, y=PTS)) +
  geom_point(color="red", size=4) +
    geom_line(color="red") +
  xlab("Season") + 
  ylab("Average Points per Game")+
  theme(axis.text.x=element_text(angle=60, hjust=1))
```

### steps:

convert to wide form to get one row per game (should be 1230 games)
compute what team won (home vs away?)
predict win with: Team_orb_pct (offensive rebounds), Team_tov_pct (turnover pct), Team_pace (straight or +/- opp pace), home vs away


Fiter to have first team listed be home (reference team), and then only one occurance so we can select just team statistics. Should be 1230 rows (games)

```{r}
nba <- transform(nba,Team_ID=as.numeric(factor(Team_Abbrev)))
nba <- transform(nba,Game_ID=as.numeric(factor(game_id)))
nba <- transform(nba,Player_ID=as.numeric(factor(player_id)))


d<- nba %>% group_by(game_id, Team_Abbrev) %>%
  mutate(Team_Ft_Attempts= sum(fta),
         Team_Assts= sum(ast), 
         Team_Steal= sum(stl),
         Team_tov=sum(tov),
         Team_asst_to_tov= Team_Assts/Team_tov)
head(d)

d2<- d %>% group_by(game_id) %>%
  filter(H_A=="H") %>%
  filter(row_number()==1) %>%
  select(game_id, H_A, Team_Abbrev, Team_Score, Team_Ft_Attempts, Team_Assts, Team_orb_pct, Team_Steal, Team_tov, Team_asst_to_tov, Opponent_Abbrev, Opponent_Score, Opponent_tov_pct, Opponent_orb_pct, Team_ID, Game_ID) %>% 
  mutate(Team_W= ifelse(Team_Score>Opponent_Score,1,0),
         Conf= ifelse(Team_Abbrev=="CHO" | Team_Abbrev=="CLE" | Team_Abbrev=="NYK" | Team_Abbrev=="ORL" | Team_Abbrev=="PHI" | Team_Abbrev=="IND" | Team_Abbrev=="CHI" | Team_Abbrev=="MIL" | Team_Abbrev=="DET" | Team_Abbrev=="ATL" | Team_Abbrev=="MIA" | Team_Abbrev=="TOR" | Team_Abbrev=="WAS" | Team_Abbrev=="BOS", "East", "West"))


nbacolors<- teamcolors %>% filter(league=="nba") %>%
  select(name, primary, secondary, tertiary, division)
nbacolors

LUT<- c("CHO" = nbacolors$name[4], 
          "CLE" = nbacolors$name[6],
        "NYK" = nbacolors$name[20],
        "ORL" = nbacolors$name[22],
        "PHI" = nbacolors$name[23],
        "IND" = nbacolors$name[12],
        "CHI" = nbacolors$name[5],
        "MIL" = nbacolors$name[17], 
        "DET" = nbacolors$name[9],
        "ATL" = nbacolors$name[1],
        "MIA" = nbacolors$name[16],
        "TOR" = nbacolors$name[28],
        "WAS" = nbacolors$name[30],
        "BOS" = nbacolors$name[2],
        "BRK" = nbacolors$name[3],
        "DAL" = nbacolors$name[7],
        "DEN" = nbacolors$name[8],
        "HOU" = nbacolors$name[11],
        "LAC" = nbacolors$name[13],
        "MEM" = nbacolors$name[15],
        "MIN" = nbacolors$name[18],
        "NOP" = nbacolors$name[19],
        "PHO" = nbacolors$name[24],
        "POR" = nbacolors$name[25],
        "SAC" = nbacolors$name[26],
        "GSW" = nbacolors$name[10],
        "SAS" = nbacolors$name[27],
        "OKC" = nbacolors$name[21],
        "LAL" = nbacolors$name[14],
        "UTA" = nbacolors$name[29])


d2$Team<- LUT[d2$Team_Abbrev]
head(d2)
head(n)

d2<- merge(d2, nbacolors, by.x = "Team", by.y = "name")
head(d2)
```

```{r}
ggplot(d2, aes(x = Team_Assts, y = Team_Abbrev, fill = primary )) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

```{r}
ggplot(d2, aes(x = Team_tov, y = Team_Abbrev, fill = Team_Abbrev)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

```{r}
ggplot(d2, aes(x = Team_Ft_Attempts, y = Team_Abbrev, fill = Team_Abbrev)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```


```{r}
ggplot(d2, aes(x = Team_orb_pct, y = Team_Abbrev, fill = d2$primary)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

# descriptives

```{r}
head(d2)
describe(d2)

```


# in mcmc

shouldn't have to add a fixed effect (?) for game because we are only predicting home team win here -- so there aren't double games

```{r}

dat <- with( d2 , list(
Team_W = Team_W,
Team_ID = Team_ID,
Team_asst_to_tov = Team_asst_to_tov,
Team_Ft_Attempts = Team_Ft_Attempts,
Team_orb_pct= Team_orb_pct,
Team_Steal= Team_Steal) )


win_nonscoring <- ulam(
alist(
Team_W ~ dbinom( 1 , p ) ,
logit(p) <- a[Team_ID] + b*( Team_asst_to_tov) + b2*(Team_Ft_Attempts)+ b3*(Team_orb_pct) +b4*(Team_Steal),
a[Team_ID] ~ dnorm( 1, 0.1 ) ,
b ~ dnorm( 3 , 1),
b2 ~ dnorm(20,5),
b3 ~ dnorm(0,1.5),
b4 ~ dnorm(6,3)
) , data=dat , chains=4 , log_lik=TRUE)

precis(win_nonscoring, depth=2)
```


### prior sim

```{r}
prior <- extract.prior( win_nonscoring , n=1e4 )
p <- inv_logit( prior$a )
dens( p , adj=0.1 )
```
our priors show most of the probability space of winning is between .7 and .8. what happens if we use flatter priors?



### sample from posterior -- win probability estimates

```{r}
post <- extract.samples(win_nonscoring)
p_left <- inv_logit( post$a )
plot( precis( as.data.frame(p_left) ) , xlim=c(0,1) )

### this is comparing team 26 (kings) to team 15 (grizzlies)
diff_a <- post$a[,26] - post$a[,15]
diff_p <- inv_logit(post$a[,26]) - inv_logit(post$a[,15])
precis( list( diff_a=diff_a , diff_p=diff_p ) )
```

outcome scale of winning for each team. are we surprised these are all so similar?

team winning percentage
```{r}
pl <- by( d2$Team_W , list( d2$Team_Abbrev) , mean )
```

## could re run that analysis and predict score differential

```{r}
dat2 <- with( d2 , list(
Team_Score = Team_Score,
Team_ID = Team_ID,
Team_asst_to_tov = Team_asst_to_tov,
Team_Ft_Attempts = Team_Ft_Attempts,
Team_orb_pct= Team_orb_pct,
Team_Steal= Team_Steal) )


points_nonscoring <- ulam(
alist(
Team_Score ~ dnorm( mu , sigma ) ,
mu <- a[Team_ID] + b*( Team_asst_to_tov) + b2*(Team_Ft_Attempts)+ b3*(Team_orb_pct) +b4*(Team_Steal),
a[Team_ID] ~ dnorm( 1, 0.1 ) ,
b ~ dnorm( 3 , 1),
b2 ~ dnorm(20,5),
b3 ~ dnorm(0,1.5),
b4 ~ dnorm(6,3),
sigma ~ dexp(1)
) , data=dat2 , chains=4 , log_lik=TRUE)

precis(points_nonscoring, depth=2)

plot(precis(points_nonscoring))


post <- extract.samples(points_nonscoring)
p_left <- inv_logit( post$a )
plot( precis( as.data.frame(p_left) ) , xlim=c(0,1) )

### this is comparing team 26 (kings) to team 15 (grizzlies)
diff_a <- post$a[,26] - post$a[,15]
diff_p <- inv_logit(post$a[,26]) - inv_logit(post$a[,15])
precis( list( diff_a=diff_a , diff_p=diff_p ) )

```


### team level analysis -- use home/away, personal stats? to predict # of points



# in bayes

here: we are allowing varying intercepts for team and varying intercepts for game just normal indexed slope for home vs away.

plot some kind of semi basic posterior prediction after this.  (if we do binary w/l it might be easier to plot like the tadpol one)

```{r}

dat_2 <- with( nba , list(
Game_ID = Game_ID,
Team_ID=Team_ID,
H_A=ifelse(H_A=="H", 1, 2),
Team_Score=Team_Score) )


points_H_A <- ulam(
alist(
Team_Score ~ dnorm( mu , sigma ) ,
mu <- a[Team_ID] + g[Game_ID] + b[H_A],
a[Team_ID] ~ dnorm( 1, 0.1 ) ,
g[Game_ID] ~ dnorm(1, 0.1),
b[H_A] ~ dnorm( 7 , 3),
sigma ~ dexp(1)
) , data=dat_2 , chains=4 , log_lik=TRUE)


tail(precis(points_H_A, depth=2))

### compute contrast

ps<- extract.samples(points_H_A)

diff_b <- ps$b[,1] - ps$b[,2] 
precis( list( diff_b=diff_b ) )


### plot
ps_d<- as.data.frame(ps)
head(ps_d)

teams<- seq(2,30,1)
dens(ps$a[,i], xlim = c(0, 2), lwd = 3, col = i, xlab = 
"posterior mean difference in team points for home vs away")
for (i in teams) {
  dens(ps$a[,i], xlim = c(0, 2), lwd = 3, col = i, xlab = 
"posterior mean difference in team points for home vs away", add=T)
}

team1 <- rnorm(1e4, ps$a[,1], ps$sigma)
teams<- seq(2,30,1)
dens(team1, lwd = 3, col = 1, xlab = 
"posterior predicted mean difference in team points for home vs away")
for (i in teams) {
  teami <- rnorm(1e4, ps$a[,i], ps$sigma)
  dens(teami, lwd = 3, col = i, add = TRUE)
}


```


last analysis: team level do individual player stats impact points
```{r}

samp<- sample(nba$game_id, 50, replace=F)
val<- subset(nba, nba$game_id %in% samp)


dat_full <- with( nba , list(
Game_ID = Game_ID,
Player_ID = Player_ID,
plus_minus = plus_minus,
double_double = ifelse(double_double==1,1,2),
drb_pct= drb_pct,
Team_Score=Team_Score) )

points_H_A <- ulam(
alist(
Team_Score ~ dnorm( mu , sigma ) ,
mu <- a[Team_ID] + g[Game_ID] + b[H_A],
a[Team_ID] ~ dnorm( 1, 0.1 ) ,
g[Game_ID] ~ dnorm(1, 0.1),
b[H_A] ~ dnorm( 7 , 3),
sigma ~ dexp(1)
) , data=dat_2 , chains=4 , log_lik=TRUE)

points_players <- ulam(
alist(
Team_Score ~ dnorm( mu , sigma ) ,
mu <- a[Game_ID] + g[Player_ID] + b*plus_minus + b2[double_double] + b3*(drb_pct),
a[Game_ID] ~ dnorm( 1, 0.1 ) ,
g[Player_ID] ~ dnorm(1, 0.1),
b ~ dnorm( 0, 3 ),
b2[double_double] ~ dnorm(0, 3),
b3 ~ dnorm(0, 3),
sigma ~ dexp(1)
) , data=dat_full, chains=4 , cores=4 , log_lik=TRUE )

tail(precis(points_players, depth=2))

post <- extract.samples(points_players)
diff_b <- post$b2[,1] - post$b2[,2]
precis( list( diff_b=diff_b ) )
```



### plot these relationships by team


```{r}
library(RColorBrewer)
myColors <- nbacolors$primary
names(myColors) <- levels(d2$Team_Abbrev)
colScale <- scale_colour_manual(name = "Team_Abbrev",values = myColors)
```


```{r}
p<- ggplot(data=d2, mapping=aes(x=Team_Ft_Attempts, y=Team_Score, colour=Team_Abbrev))
p + geom_point(color = d2$primary) + geom_smooth(method = "lm", se=F)+
  facet_grid(~division)+
  labs(x="Number of Team Free Throw Attempts per Game", y="Total Team Points per Game") +
   scale_color_manual(values=d2$primary)+
  colScale
```

```{r}
p<- ggplot(data=d2, mapping=aes(x=Team_asst_to_tov, y=Team_Score, colour=Team_Abbrev))
p + geom_point(color=d2$primary) + geom_smooth(method = "lm", se=F)+
  facet_grid(~division)+
  labs(x="Number of Team Assists to Turnovers Ratio per Game", y="Total Team Points per Game")+
  scale_color_manual(values=d2$primary)+
  colScale

```


```{r}
p<- ggplot(data=d2, mapping=aes(x=Team_orb_pct, y=Team_Score, colour=Team_Abbrev))
p + geom_point(color = d2$primary) + geom_smooth(method = "lm", se=F)+
  facet_grid(~division)+
  labs(x="Team Offensive Rebound Percentage per Game", y="Total Team Points per Game")+
  colScale
```



```{r}
p<- ggplot(data=d2, mapping=aes(x=Team_Steal, y=Team_Score, colour=Team_Abbrev))

p + geom_point(color = d2$primary) + geom_smooth(method = "lm", se=F)+
  facet_grid(~division)+
  labs(x="Number of Team Steals per Game", y="Total Team Points per Game", fill="Team") +
  colScale +
   theme(legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(1, 'cm'), #change legend key height
        legend.key.width = unit(1, 'cm'), #change legend key width
        legend.title = element_text(size=14), #change legend title font size
        legend.text = element_text(size=10)) #change legend text font size

```

